%
% $Id: chapter6.tex 2628 2008-06-04 11:21:59Z jozinke $
%
%%\pagestyle{scrheadings}
%%\ohead[]{}
%%\ihead[]{}
%%\chead[]{}
%%\ofoot[\pagemark]{\pagemark}
%%\ifoot[]{}
\chapter{Conclusion}

In this thesis, a new, shared memory based \ac{IPC} architecture for the transmission of log messages was developed.
Additionally, the proof of concept \ac{IPS} Simplefail2ban was implemented and evaluated, in conjunction with the \ac{IPC} architecture.
Overall, the evaluation results are generally positive. Both the  logfile and shared memory based variants of Simplefail2ban largely performed as expected, blocking more than 95\% of all unwanted
traffic, in all but two measurement scenarios. Performance differences between logfile and shared memory
based transmission of logfiles, where surprisingly small, with the exception of the 10 million \ac{PPS} measurement.
Compared to Fail2ban, significantly higher loads, both in terms of unwanted traffic 
rate a well as the number of clients, were achieved. However, issues with the replication
of the measurement for the 10 million \ac{PPS} scenario, present a significant caveat to the results. The
evaluation for Simplefail2ban under high traffic loads, will have to be thoroughly replicated, to obtain a more 
realistic estimate of the average performed. Unfortunately, this could not be done in time for submission of this thesis
and will be handed in as an addendum. 
\par
While the performance ceiling of Simplefail2ban remains to be more accurately assessed, 
the minimal goal of outperforming Fail2ban for the \ac{DoS} scenario, described in \ref{sec:fail2ban}
was achieved. Therefore, this thesis demonstrated, that the shared memory based transmission of log messages, can
deliver better performance, than the traditional logfile-parsing based approach.
\par 
For future development of the shared memory architecture, the focus could be, to 
improved the integrability of the design with real world applications. The current \ac{API}
only offers a very low level interface, that could be improved, to support a standardizes protocol like syslog, or possibly
plugins for log aggregators such as Logstash. As aforementioned, the performance of Simplefail2ban also needs
to be further evaluated, both in regards to its behavior under high load, as well as to the overwrite feature, which
did not perform as well as expected. Additionally, further implementation of \ac{IPC} based based logging systems could be added
and be compared to the performance of the shared memory architecture.
